---
title: "HDGCvar Vignette"
author: "Luca Margaritella"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HDGCvar Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# HDGCvar

<!-- badges: start -->
[![CRAN_Version_Badge](http://www.r-pkg.org/badges/version/HDGCvar)](https://cran.r-project.org/package=HDGCvar)
[![CRAN_Downloads_Badge](https://cranlogs.r-pkg.org/badges/grand-total/HDGCvar)](https://cran.r-project.org/package=HDGCvar)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

`HDGCvar` allows for testing Granger causality in High Dimensional Vector Autoregressive Models (VARs). 
Granger causality can be tested between time series that are stationary (`HDGC_VAR_I0`), non stationary (unit root), cointegrated, or all the above (`HDGC_VAR`). Bivariate as well as multivariate (i.e. blocks) causality can be considered by specifying the name(s) of the variable(s) of interest in `GCpair` (or `GCpairs`) and networks can be plotted to visualize the causal structure among several variables using `Plot_GC_all`.

A specific part of `HDGCvar` is dedicated to Realized Volatilities (RV), thus using the Heterogeneous VAR (HVARs). It gives the possibiity of building RV spillover networks (`HDGC_HVAR_all`) as well as conditioning RV on Realized Correlations (`HDGC_HVAR_RV_RCoV_all`).



## Installation

You can install the released version of HDGCvar from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("HDGCvar")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Marga8/HDGCvar")
```


All the functions in `HDGCvar` are based on the following two papers: 

1. A. Hecq, L. Margaritella, S.Smeekes, "Granger Causality Testing in High Dimensional VARs: a Post Double Selection Procedure"(2019, https://arxiv.org/abs/1902.10991 )  
2. A. Hecq, L. Margaritella, S.Smeekes, "Inference in Non Stationary High Dimensional VARs" (2020, check the latest version at https://sites.google.com/view/luca-margaritella ).


## Details
If the addition of variable X to the given information set &Omega; alters the conditional distribution of another variable Y, and both X and &Omega; are observed prior to Y, then X improves predictability of Y, and is said to *Granger cause* Y with respect to &Omega;.

Statistically assessing the predictability among two (or blocks of) time series turns out to be a fundamental concept in modern time series analysis. Its applications range over from macroeconomics, finance, network theory and even the neurosciences.

With the increased availability of larger datasets, these causality concepts have been extended to the high-dimensional setting where they can benefit from the inclusion of many more series within the available information set &Omega;.
Note how including as much information as possible is particularly tending to the original idea of Granger (1969) who envisioned &Omega; to be "all the information in the universe", in order to avoid spurious causality.

Conditioning on so many variables comes at a cost, i.e. quickly running into problems of high-dimensionality: sky-rocketing variance, overfitting and in general invalidity of standard statistical techniques.

A. Hecq, L. Margaritella, S.Smeekes (2019), designed a Granger causality LM test for high-dimensional stationary vector autoregressive models (VAR) which combine dimensionality reduction techniques based on penalized regressions such as the lasso of Tibshirani (1996), with the post-double-selection procedure of Belloni et al. (2011) to select the set of relevant covariates in the system. The double selection step allows to substantially reduce the omitted variable bias and thereby allowing for valid post-selection inference on the parameters. 

*If you are sure that your time series are stationary*, you can go ahead and use one of the following functions from `HDGCvar`:  
  
1. `HDGC_VAR_I0` which tests for Granger causality in High Dimensional Stationary VARs  
2. `HDGC_VAR_multiple_I0` or `HDGC_VAR_multiple_pairs_I0` which test multiple combinations Granger causality in High Dimensional Stationary VARs  
3. `HDGC_VAR_all_I0` which allows you to test all the bivariate combinations in you dataset and hence to build a Granger causality network.  

All these functions ask you as inputs:

* `GCpair` or `GCpairs`, a list object explicitly containing the Granger caused variables "GCto" and the Granger causing "GCfrom".  
* `data` , the dataset containing as columns all the time series you are modeling.  
* `p = 1`, the lag-length of the VAR. Unless you want to explicitly impose it, you can use `HDGCvar::lags_upbound_BIC()` which estimates an empirical upper bound for the lag-length using Bayesian Information Criteria (for details see Hecq et al. 2019).  
* `bound = 0.5 * nrow(data)`, this is meant for applying a lower bound on the penalty parameter of the lasso. In simple words it means you are telling the lasso, in each equations it estimates, to please stop when it has reached an amount of selected (i.e. non-zero coefficient) variables corresponding to $50\%$ of the sample size. This is innocuous in small systems but quite important for systems where the number of variables per-equation (i.e. lags included) is larger than the sample size (for details see Hecq et al. 2019).  
* `parallel = FALSE`, put equal to TRUE for parallel computing.  
* `n_cores = NULL`, number of cores to use in parallel computing, default is all but one.  


A. Hecq, L. Margaritella, S.Smeekes (2020) extended the Post-double-selection Granger causality LM test of Hecq et al. (2019) to the case in which your system might contain time series integrated of arbitrary orders and possibly even cointegrated. To accomplish this is necessary to simply augment the lags of the variables of interest by the maximum order of integration we suspect the series having. This allows to completely avoid any biased pre-test of unit-root or cointegration and directly perform high-dimensional inference on the desired parameters.  

*Therefore: if you are NOT sure whether your time series are stationary, non-stationary, cointegrated, or an appropriate mix of the above or you do not trust the biased unit root and cointegration tests out there*, we got you covered! you can go ahead and use one of the following functions from `HDGCvar`:  

1. `HDGC_VAR` which tests for Granger causality in High Dimensional Stationary/Non-Stationary/cointegrated or a mix of the above, VARs   
2. `HDGC_VAR_multiple` or `HDGC_VAR_multiple_pairs` which test multiple combinations Granger causality in High Dimensional Stationary/Non-Stationary/cointegrated or a mix of the above, VARs   
3. `HDGC_VAR_all` which allows you to test all the bivariate combinations in you dataset and hence to build a Granger causality network.  


All these functions ask you the same inputs as reported above for `HDGC_VAR_I0` with the addition of the following:  

* `d=2`, is the augmentation needed to handle the potentially non-stationary time series. It should correspond to the maximum suspected order of integration of your time series. This might sound vague but think about it: in economics I have never seen a variable integrated of order three. Therefore, unless you are absolutely sure that your time series are at most integrated of order one or that for whatever unlikely reason you have at least one variable integrated of order three, I would recommend to put `d=2`.  



Back to A. Hecq, L. Margaritella, S.Smeekes (2019), in their empirical application they considered a specific case of stationary time series, namely they investigate the volatility transmission in stock return prices using the daily realized variances. These are very attractive measures among practitioners and academics for modelling time varying volatilities and monitoring financial risk. Given the time series of realized volatilities, a multivariate version of the heterogeneous autoregressive model (HVAR) of Corsi (2009) is employed to model their joint behavior.  

*Therefore: if you are interested in testing Granger causality among realized volatilities and potentially build volatility spillover networks* you should use one of the following function from `HDGCvar`:  

1. `HDGC_HVAR` which tests for Granger causality in High Dimensional stationary HVARs.   
2. `HDGC_HVAR_multiple` or `HDGC_VAR_multiple_pairs` which test multiple combinations Granger causality in High Dimensional stationary HVARs. 
3. `HDGC_HVAR_all` which allows you to test all the bivariate combinations in you dataset and hence to build a Granger causality network.  

All these functions ask you the same inputs as reported above for `HDGC_VAR_I0` with the inclusion of `log=TRUE` to log-transform the series (recommended) and the exclusion of the lag-length parameter `p` as in a HVAR this is already pre-specified to be equal to 3, namely the daily, weekly, monthly aggregates of realized volatilities (for details see Corsi (2009) and Hecq et al. (2019)).

Note: to account for the potential heteroskedasticity in the data, all these functions returns the Asymptotic &chi;$^2$ version of the LM test ("Asymp"), the small sample correction F tests ("FS_cor") and the heteroskedasticity-robust Asymptotic &chi;$^2$ (Asymp_Robust).  

It is not all: if you whish to condition your volatility spillover network on realized correlations you can also do so by using one of the following function from `HDGCvar`:  

1. `HDGC_HVAR_RV_RCoV_all`, compute Network of Realized Volatilities conditional on the set of Realized Correlations  
2. `HDGC_HVAR_RVCOV`, Test Granger causality for Realized Volatilities in High Dimensional Stationary Heterogeneous VARs conditioning on Realized Correlations    
3. `HDGC_HVAR_multiple_RVCOV`, Test multiple combinations Granger causality for realized volatilities in High Dimensional Stationary HVARs  

All these functions ask you the following inputs:

* `realized_variances`, matrix of realized variances.  
* `realized_covariances`, matrix of realized covariances.  
* `fisher_transf=TRUE`, to Fisher-transform the realized correlations. It essentially calls the function `HDGCvar::Realized_corr()` which calculates the realized correlations from the inputs `realized_variances` and `realized_covariances` and Fisher-transform them (for details see Hecq et al. (2019)).  
* `log=TRUE`, to log-transform the volatility series (recommended).   
* `bound = 0.5 * nrow(realized_variances)`, same as `bound = 0.5 * nrow(data)`, see `HDGC_VAR_I0`.
* `parallel = FALSE`, put equal to TRUE for parallel computing.     
* `n_cores = NULL`, number of cores to use in parallel computing, default is all but one.     

In order to plot networks created via: `HDGC_VAR_all_I0`,  `HDGC_VAR_all`, `HDGC_HVAR_all`,  `HDGC_HVAR_RV_RCoV_all` you can use the function  

* `Plot_GC_all`  

This function takes as inputs:  
* `Comb`, stands for the result of one of the functions: `HDGC_VAR_all_I0`,  `HDGC_VAR_all`, `HDGC_HVAR_all`,  `HDGC_HVAR_RV_RCoV_all`  
* `Stat_type="FS_cor"`, the default is "FS_cor" namely the small sample correction F test. Alternatively one can use "Asymp" for the asymptotic &chi;$^2$ test or "Asymp_Robust" for the heteroskedasticity-robust asymptotic &chi;$^2$ test.  
* `alpha=0.01`, the desired probability of type one error, default is 0.01.  
* `multip_corr`, a list A list: first element is logical, if TRUE a multiple testing correction using stats::p.adjust() is used. The second element of the list define the p.adjust.method=c("holm", "hochberg", "hommel", "bonferroni", "BH", "BY","fdr", "none")). If the second element gets the name "APF_FDR" then APFr::apf_fdr which uses empirical Bayes is called and a third and fourth elements in the mutip_corr list are required: gamm=c(,,) requires a min, max and step length values to be set for the threshold on the p_values,fdr.apf=c(,) requires one or two values: either (NULL,value) or (value,NULL) if one wants to have specified amount of average power (fdr) no matter fdr (average power).
If both (value,value) are given, the calculated threshold will find the closest combination to both apf and fdr desired. The last element of the list is logical: verbose=TRUE if one wants to know how much apf/fdr the testing has.  
* `...`, all parameters for the network plot: see example and igraph documentation.  
* `cluster`, a list: first element is logical, if TRUE a cluster plot using igraph::cluster_edge_betweenness() is plotted. Other elements are respectively: vertex.size, vertex.label.color,vertex.label.cex, vertex.label.dist, edge.curved (see igraph for details).


## Examples

Here is a simulated examples which show you how to use some of the features in `HDGCvar`.

```{r example, echo=TRUE, fig.height=5, fig.width=5,dpi=130}
library(HDGCvar)
library(igraph)

#First, let us simulate some time series
#The SimulVAR below simulates a (non-sparse) stationary VAR(1) with sample size T_ and number of variables g
SimulVAR=function(T_,g){
  coef1<-matrix(NA,nrow=g,ncol=g)
  for (i in 1:g) {
    for (j in 1:g) {
      coef1[i,j]<-((-1)^(abs(i-j)))*(0.4^(abs(i-j)+1))
    }
  }
  presample<-1
  T_new<-T_+presample
  eps1<-rnorm(ncol(coef1)*T_new,0,1)
  eps<-matrix(eps1,nrow=ncol(coef1))
  X <- matrix(nrow=ncol(coef1),ncol=T_new)
  X[,1] <- eps[,1]
  for (t in 2:T_new) {
    X[,t] <- (coef1)%*%X[,t-1]+eps[,t]
  }
  finseries<- X[,(1+presample):T_new]
  return(t(finseries))
}

#You can check the stationarity of the VAR(1) by simply observing:
g=20
coef<-matrix(NA,nrow=g,ncol=g)
  for (i in 1:g) {
    for (j in 1:g) {
      coef[i,j]<-((-1)^(abs(i-j)))*(0.4^(abs(i-j)+1))
    }
  }
max(eigen(coef)$values) #<1

#Let us fix some seed and simulate 20 non-stationary time series with sample size 100 from a VAR(2) by taking the inverse difference of the stationary series simulated
set.seed(123)
dataset_I0<-as.matrix(SimulVAR(100,20))
dataset<-as.matrix(diffinv(dataset_I0))
colnames(dataset)<-c(paste(rep("Var",20),1:20))

#Let us plot the time series to get a feeling of what kind of data we are dealing with
ts.plot(dataset)

#Select the lag-length of the VAR using HDGCvar::lags_upbound_BIC
selected_lag<-lags_upbound_BIC(dataset,p_max=10)
print(selected_lag)
#Note: hooray! the selected lag is p=2 which it is correct given our simulated data! Of course in practice we could not know this.

#Suppose we are interested in testing whether variable name "Var5" Granger-causes variable name "Var1" given all other variables in your dataset.
interest_variables=list("GCto"="Var 1","GCfrom"="Var 5")

#By visual diagnostics on the series in our dataset from the previous plot we definitely have some time series with unit roots and probably cointegration. Actually: we definitely know this as we created the process in such a way that all the variables have a unit root, in practice obviously this would not be the case.

#However, the nice feature of HDGCvar is that you can completely avoid any pre-tests of integration/cointegration of your series. As we do not know whether the maximum order of integration of our time series is one or two, the safe choice is to let d=2 and simply run

HDGC_VAR(GCpair=interest_variables, data=dataset, p = selected_lag, d = 2, bound = 0.5 * nrow(dataset),
                     parallel = T, n_cores = NULL) 

#Now, suppose we are interested in testing multiple Granger causality relations
mult_interest_variables<-list(list("GCto"="Var 7", "GCfrom"="Var 19"),list("GCto"="Var 4", "GCfrom"="Var 16"))
HDGC_VAR_multiple(dataset, GCpairs=mult_interest_variables, p= selected_lag, d=2, bound = 0.5 * nrow(dataset),
                     parallel = T, n_cores = NULL)


#Let us now estimate the full network of causality among the 20 series in dataset
network<-HDGC_VAR_all(dataset, p = selected_lag, d = 2, bound = 0.5 * nrow(dataset), 
                         parallel = TRUE, n_cores = NULL)

#Now let us plot the estimated network.
#You can even do clustering of the connected series using the option cluster=list(T,5,"black",0.8,1,0).
#This uses Newman-Girvan (2002) algorithm based on edge-betweenness.
Plot_GC_all(network, Stat_type="FS_cor",alpha=0.01, multip_corr=list(F),directed=T, layout=layout.circle, main="Network",edge.arrow.size=.2,vertex.size=5, vertex.color=c("lightblue"), vertex.frame.color="blue",vertex.label.size=2,vertex.label.color="black",vertex.label.cex=0.6, vertex.label.dist=1, edge.curved=0,cluster=list(T,5,"black",0.8,1,0)) 
```
